---
title: "R Tutorial"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
data = read.csv("https://raw.githubusercontent.com/dbowden/Rtutorials/master/regressionlab_condensed.csv")
```


## Introduction

### R

R is a programming language for managing data, making graphs, and performing statistical analyses. Don't worry! It isn't as hard as it sounds, and you won't be tested on this on the midterm. This tutorial is meant to give you a feel for what it would be like to use quantitative methods in your thesis, and to provide illustrations for some of the concepts from the lectures.

### The Data

We will use the same example for the remainder of our quantitative unit. Many scholars assert that democracies tend to have stronger economic growth than non-democracies. We will test this proposition using data on regime scores from the [Polity Project](https://www.systemicpeace.org/inscrdata.html) and economic data from the [World Bank](https://data.worldbank.org/indicator/NY.GDP.MKTP.CD). For simplicity, we will use the average values of these measures for the years 2008-2017, and we will use each of the 169 countries for which there is data on these measures.

### Basic Math 

R requires that you type commands into a console rather than clicking a button. One type of command is simple math. You can use R just like a calculator. Try typing `1+2` into the 'Code' box below and then hit the 'Run Code' button.

```{r two-plus-two, exercise=TRUE, exercise.lines = 1}

```

A new box should appear with the answer. There are many other operations available including multiplication (`*`), division (`/`) and exponents (`^`). Try a few math problems before moving on.

## Functions

Thankfully, we rarely need to type out the math by hand. Instead, we can rely mostly on built-in commands called functions. A function is a word followed by parentheses that contain options such as the dataset to be used with the function. One of the simplest functions is called `print`. It does exactly what you would expect - it prints the message you supply as an option in the parentheses. Hit 'Run Code' to see for yourself.

```{r hello, exercise=TRUE, exercise.lines = 1}
print("Hello world!")
```

Note that in R all text (except the words `TRUE` and `FALSE` and their abbreviations `T` and `F`) must be surrounded by quotation marks. You can use double or single quotation marks, but each open quotation mark MUST be followed by a close quotation. Let's see what happens if we fail to do that:

```{r hello-error, exercise=TRUE, exercise.lines = 1}
print("Hello world!)
```

You should see a confusing error message. See if you can fix the command so that it works properly.

Also note that R commands are case-sensitive. See what happens if we capitalize the "P" in `print`:

```{r hello-case, exercise=TRUE, exercise.lines = 1}
Print("Hello world!")
```

R's perfectionist tendencies like these are some of the most common sources of struggle for new users. Don't be afraid to ask for help! Even experienced users sometimes struggle to catch typos.

## Loading Data

Let's move onto a more useful function. The data mentioned at the beginning is stored in a .csv file (short for comma-separated values, a type of spreadsheet file). We can open in R using the `read.csv` command with the filename or url as the option. On its own, `read.csv` will simply print the contents of the file in the results window. To save the data for use with other commands, we need to give it a name. We can use any name we want so long as it begins with a letter, and we can use `=` or `<-` to store the data under said name. The `head` command will show us the first few rows of data. Click 'Run Code' below.

```{r readcsv, exercise=TRUE, exercise.lines = 2}
data = read.csv("https://raw.githubusercontent.com/dbowden/Rtutorials/master/regressionlab_condensed.csv")
head(data)
```

As you can see, we have a country name, `gdpgr`, the average GDP growth rate (as a % change from previous year) and `polity2`, the average Polity score (a scale ranging from -10 [most authoritarian] to +10 [most democratic]) for the period 2008-2017. We also have a binary indicator called `democ` with a value of 1 for states with an average Polity score of 6 or higher, and 0 for all others.

### Exploring the Data

We can obtain some basic information about the dataset and variables by using the summary command:

```{r summary, exercise=TRUE, exercise.lines = 1}
summary(data)
```

The 'Length' attribute of the country variable tells us that we have 169 observations. That means that we are missing data on about 25 UN members, mostly because Polity does not provide data on states with populations below a certain threshold. We can also see the the mean, median, and quartile values for the other variables. 'NA' is the code R uses to represent missing values. Notice that there are a few missing values on GDP growth.

## Graphs

### Histograms

It is easy to make graphs in R. A useful type of graph for examining the distribution of our variables is a histogram. The command for making on is called `hist`. As an argument we will supply the name of our dataset, and we can use the `$` symbol to specify which variable in the dataset we wish to graph. We can make a histogram of GDP growth using this command:

```{r hist1, exercise=TRUE, exercise.lines = 1}
hist(data$gdpgr)
```

This is a fairly normal, bell curve-like distribution

Now try making your own histogram of the average polity score (the `polity2` variable). How would you describe this distribution?

```{r hist2, exercise=TRUE, exercise.lines = 1}

```

### Scatterplots

A scatterplot is often a useful graph for looking at the relationship between two variables. These are very simple to make using the `plot` command with the two variables of interest:

```{r scatter, exercise=TRUE, exercise.lines = 1}
plot(data$polity2, data$gdpgr)
```

Unfortunately in this case the relationship is not totally clear from the scatterplot.

### Box Plot

Creating a boxplot with the binary `democ` variable might be more revealing. The options are a bit different here - we need to supply the variable we are plotting separated from the grouping variable with a `~` symbol:

```{r boxplot, exercise=TRUE, exercise.lines = 1}
boxplot(gdpgr ~ democ, data = data)
```

The plot suggests that democracies have *lower* economic growth than non-democracies! In the next section we'll check the statistical significance of this pattern.

## T-Tests

### Subsets and Means

First, let's calculate the average growth for each group. We can wrap the `with` command around the `mean` command to use a subset of the data. Note that we need to set `na.rm = TRUE` to remove missing values. Otherwise, R will give us a missing value as the mean.

```{r mean-democ, exercise=TRUE, exercise.lines = 2}
print("Mean GDP growth for democracies:")
with(subset(data, democ == 1), mean(gdpgr, na.rm = TRUE))
```

```{r mean-nondemoc, exercise=TRUE, exercise.lines = 2}
print("Mean GDP growth for non-democracies:")
with(subset(data, democ == 0), mean(gdpgr, na.rm = TRUE))
```

Consistent with the boxplot, we can see that non-democracies have a higher average growth rate than democracies. That is a very interesting result, but we shouldn't jump to conclusions just yet. As always, we should think of our data as a sample, and use a statistical test to determine whether we might expect to see the same pattern in other samples or the full population.

### T-Tests

A two-sample t-test is commonly used to assess whether the difference between two groups is statistically significant. It essentially asks whether the difference is large enough and/or consistent enough that it is unlikely to be the result of random chance. The command is similar to the boxplot - we specify a dependent variable separated from a group variable.

```{r ttest, exercise=TRUE, exercise.lines = 1}
t.test(gdpgr ~ democ, data = data)
```

What is the p-value? Is it small enough for this difference between democracies and non-democracies to be statistically significant?

Do you see any limitations to this type of analysis?